---
title: "Python Configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```


survivalGPU is a package based on a package written in python, dependant on the `reticulate` R package. To use it, it's necessary to have installed `torch`, `torch-scatter`, and `pykeops` packages in your python environment. Moreover, it's possible to check through `reticulate` the python executable, and to check if the packages mentioned above are installed.


```{r setup}
library(reticulate)
```


```{r}
# Check your python executable
py_config()
py_exe()

# Check installed packages in your python environment
py_list_packages()
```

Attention, if you don't have install the `reticluate` R package, install it (with `install.packages("reticulate")`). During installation, R will offer you to install miniconda. If you accept, be careful because your python executable & pyhton home will change for your new miniconda environment by default, without torch, torch-scatter and pykeops. You can install them in miniconda, or you can change your python environment after starting a new R session. I recommend to not install miniconda to use the python environment on which we are used to working.

Check the [`reticulate` R documentation](https://rstudio.github.io/reticulate/articles/versions.html) to see other options to manage your python configuration.

```{r}
# Test survivalGPU import, and the use of CUDA for GPU support
library(survivalGPU)
use_cuda()
```
