---
title: "Python Configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```


survivalGPU relies on a Python backend, through the `reticulate` R package. It needs a Python environment (Python >= 3.10) with a few libraries installed, such as `torch`. `pykeops` is only required for `wceGPU()` (the WCE model): it is not available on Windows, where `coxphGPU()` works without it. To use `wceGPU()` on Windows, install R inside [WSL2](https://learn.microsoft.com/windows/wsl/install) (the Windows Subsystem for Linux) and follow the Linux instructions.


## Install the Python libraries

Create a Python environment called `survivalGPU`, and install the libraries in it. You only do this once; `torch` is a large download, so it takes a few minutes.

```{r}
library(reticulate)

virtualenv_create("survivalGPU", version = ">=3.10")
virtualenv_install("survivalGPU", c("numpy", "torch", "pandas", "scipy",
                                    "matplotlib", "beartype", "jaxtyping"))

# Only for the WCE model, wceGPU() -- not available on Windows:
virtualenv_install("survivalGPU", "pykeops")
```

On macOS, `pykeops` also needs OpenMP: run `brew install libomp` in a terminal.

If `virtualenv_create()` fails with "Suitable Python installation for creating a venv not found", the Python it picked can't create virtual environments (e.g. on Ubuntu, without the `python3.X-venv` package). Give it another Python explicitly:

```{r}
virtualenv_create("survivalGPU", python = "/usr/bin/python3.10")
# or, if you don't have a Python >= 3.10:
virtualenv_create("survivalGPU", python = install_python("3.11"))
```


## Use it

At the start of each R session, select the Python environment **before** loading survivalGPU:

```{r}
library(reticulate)
use_virtualenv("survivalGPU")

library(survivalGPU)
use_cuda() # TRUE if a GPU is detected
```

You can also use your own Python environment, selected with `reticulate::use_virtualenv()`, `reticulate::use_condaenv()` or the `RETICULATE_PYTHON` environment variable (e.g. in your `.Renviron` file). Check that the right Python is used with:

```{r}
py_config()
```

If R offers to install Miniconda, be careful: your Python executable would then change to that new Miniconda environment.

Check the [`reticulate` documentation](https://rstudio.github.io/reticulate/articles/versions.html) for other ways to manage your Python configuration.
