---
title: "Python Configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```


survivalGPU is a package based on a package written in python, dependant on the `reticulate` R package. To use it, it's necessary to have installed some python libraries such as `torch`, `torch-scatter`, and `pykeops` in your python environment.

Maybe it's installed in your python default system, but it's better if your create your own virtualenv for your project. Let's make the steps to configure properly a virtual python envrironment to use `survivalGPU` in R.

```{r}
library(reticulate)

# you create the 'survivalGPU' environment
virtualenv_create("survivalGPU")
virtualenv_list()
# you can check python libraries in this python environment
py_list_packages(envname = "survivalGPU") # numpy by default

# now, install all the survivalgpu python dependencies
virtualenv_install("survivalGPU", packages = c("torch", "torch_scatter",
                                               "pykeops", "matplotlib",
                                               "beartype", "jaxtyping"))
py_list_packages(envname = "survivalGPU")
```


```{r}
#  /!\ Restart your R session  /!\
library(reticulate)

# Check if it's the correct python executable
py_discover_config()
use_virtualenv(virtualenv = "survivalGPU")
py_config()
```


As you can see, you're connect at your survivalGPU python environment. So now, if you load `survivalGPU` R package after exec `reticulate::use_virtualenv(virtualenv = "survivalGPU")`, the package will use this python environment.  


Attention, if you don't have install the `reticluate` R package, install it (with `install.packages("reticulate")`). During installation, R will offer you to install miniconda. If you accept, be careful because your python executable & python home will change for your new miniconda environment. You can change your python environment after starting a new R session.  

Check the [`reticulate` R documentation](https://rstudio.github.io/reticulate/articles/versions.html) to see other options to manage your python configuration.
